---
title: "Calculate environmental variables"
author: "Ward Langeraert"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
editor_options: 
  chunk_output_type: console
---

# Goal

Calculate and add environmental variables to observations.

```{r}
#| warning: false
#| message: false
# Packages
library(tidyverse)
library(sf)

# Conflicts
conflicted::conflicts_prefer(dplyr::filter)

# Paths
data_path <- here::here("data", "raw")
interim_path <- here::here("data", "intermediate")
out_path <- here::here("data", "processed")

# Global variables
proj_crs <- 31370
```

# Select heather habitat

We load and select Natura 2000 heather habitats.

```{r}
# Load BWK layer
bwk_flanders <- st_read(file.path(data_path, "BwkHab.shp"))

# Heather starts with 4
heather_hab <- bwk_flanders %>%
  filter(grepl("^4", HAB1))
```

## Categorise habitats

Which habitat types do we get?

```{r}
table(heather_hab$HAB1)
```

<!-- spell-check: ignore:start -->

We see:

- 4010: Northern Atlantic wet heaths with *Erica tetralix*
- 4030: European dry heaths
- 4010, 4030: mix of above
- 4010,gh/4030,gh: gh = no habitat
- 4010,rbbsm: rbbsm = *Myrica gale* bush
- 4030,bos: in combination with forest

We make the following classification

- wet heaths
  - HAB1 = 4010|4010,gh|4010,rbbsm
  - IF PHAB1 >= 60
- dry heaths
  - HAB1 = 4030|4010,gh|4030,bos
  - IF PHAB1 >= 60
- mixed heaths
  - HAB1 = 4010, 4030
  - OR HAB1 = 4010|4030 and HAB2 = 4030|4010 with PHAB1 + PHAB2 >= 70 AND PHAB2 >= 20

<!-- spell-check: ignore:end -->

Other polygons contain not enough habitat and are removed.

```{r}
heather_hab_cat <- heather_hab %>%
  mutate(
    habClass = case_when(
      grepl("4010,4030", HAB1) ~ "mixed heath",
      (grepl("^4010", HAB1) & grepl("^4030", HAB2)) &
        (PHAB1 + PHAB2 >= 70 & PHAB2 >= 20) ~ "mixed heath",
      (grepl("^4030", HAB1) & grepl("^4010", HAB2)) &
        (PHAB1 + PHAB2 >= 70 & PHAB2 >= 20) ~ "mixed heath",
      grepl("^4010", HAB1) & PHAB1 >= 60 ~ "wet heath",
      grepl("^4030", HAB1) & PHAB1 >= 60 ~ "dry heath",
      TRUE ~ "invalid"
    )
  ) %>%
  filter(habClass != "invalid")

table(heather_hab_cat$habClass)
```

## Join polygons

We join neighbouring polygons together if they belong to the same habitat class.
We calculate surface area of each polygon.
We only keep polygons larger than 10,000 $m^2$.

```{r}
#| warning: false
heather_hab_cat_merged <- heather_hab_cat %>%
  group_by(habClass) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_cast("POLYGON") %>%
  mutate(habID = row_number(),
         habAreaM2 = units::drop_units(st_area(geometry))) %>%
  filter(habAreaM2 > 10000) %>%
  select(habID, habClass, habAreaM2, everything()) %>%
  st_transform(proj_crs)
```

```{r}
mapview::mapview(heather_hab_cat_merged, zcol = "habClass",
                 layer.name = "Habitat")
```

## Export layer

We write the layer out.

```{r}
st_write(
  heather_hab_cat_merged,
  file.path(interim_path, "heather_flanders_cat.gpkg"),
  append = FALSE
)
```

# Filter observation data

Load observation data of pollinator groups.

```{r}
poll_groups <- c("bees", "butterflies", "moths", "hoverflies")

# Read the files and build a named list
pollinator_obs_df <- lapply(
  file.path(interim_path, paste0(poll_groups, "_interim_data.csv")),
  read_csv,
  show_col_types = FALSE
)
names(pollinator_obs_df) <- poll_groups

# Create sf objects
pollinator_obs_sf <- lapply(
  pollinator_obs_df,
  st_as_sf,
  coords = c("verbatimLongitude", "verbatimLatitude"),
  crs = proj_crs,
  remove = FALSE
)
```

## Habitat selection

We select observations within heather habitats.

```{r}
#| warning: false
pollinator_obs_filtered <- lapply(
  pollinator_obs_sf,
  st_intersection,
  y = heather_hab_cat_merged
)
```

How many observations are retained?

```{r}
sapply(pollinator_obs_filtered, nrow)
```

Consider each observation as presence (do not use `individualCount`).
Count presences per habitat polygon, all other are absences (pseudo-absences).
We only keep species that occur in at least 10 locations.

```{r}
pollinator_obs_wide <- lapply(pollinator_obs_filtered, function(df) {
  df %>%
    st_drop_geometry() %>%
    count(habID, vernacularName) %>%
    # Select common species
    mutate(n_loc = n_distinct(habID), .by = "vernacularName") %>%
    filter(n_loc >= 10) %>%
    select(-n_loc) %>%
    # Wide format and pseudo-absences
    pivot_wider(names_from = vernacularName, values_from = n,
                values_fill = 0) %>%
    arrange(habID)
})
```

How many observations are retained?
Respectively number of sites and number of species.

```{r}
sapply(pollinator_obs_wide, dim)
```

Over all pollinator groups:

```{r}
pollinator_total_wide <- lapply(pollinator_obs_filtered, st_drop_geometry) %>%
  bind_rows() %>%
  count(habID, vernacularName) %>%
  # Select common species
  mutate(n_loc = n_distinct(habID), .by = "vernacularName") %>%
  filter(n_loc >= 10) %>%
  select(-n_loc) %>%
  # Wide format and pseudo-absences
  pivot_wider(names_from = vernacularName, values_from = n,
              values_fill = 0) %>%
  arrange(habID)
```

In total we have `r nrow(pollinator_total_wide)` sites and `r ncol(pollinator_total_wide) - 1` species.

## Write out observation dataset

Per pollinator group:

```{r}
for (i in names(pollinator_obs_wide)) {
  # Create file path
  file_name <- paste0(i, "_data_wide.csv")
  path_full <- file.path(out_path, file_name)

  # Write out dataset as csv
  write_csv(
    pollinator_obs_wide[[i]],
    path_full
  )
  print(paste("Dataset written to", path_full))
}
```

Total:

```{r}
write_csv(pollinator_total_wide,
          file.path(out_path, "pollinator_data_wide.csv"))
```

# Environmental variables

We have the following layers and variables:

- `BwkHab.shp`: habitat mapping layer of Flanders
  - heather habitat classes (see above)
  - size of habitat parcels (see above)
- `klimaatvl.shp`: climate layer with multiple variables
  - ...
- `WRB Soil Units 40k (reference soil group).lyr`
  - soil types
  
We calculate environmental variables per site at the centroid of the habitat polygon.

## Habitat layer

See above.
Brief overview here.

```{r}
site_centroids <- heather_hab_cat_merged %>%
  filter(habID %in% pollinator_total_wide$habID) %>%
  mutate(geometry = st_centroid(geometry))
```

### Habitat classes heather

We look at the number of observations per habitat class.

```{r}
site_centroids %>%
  ggplot(aes(x = habClass)) +
  geom_bar()
```

### Habitat parcel size

We have a good mix of parcel surface area for each habitat class.

```{r}
site_centroids %>%
  ggplot(aes(y = log(habAreaM2), x = habClass)) +
  geom_boxplot() +
  labs(x = "") +
  theme_minimal()
```

## Climate layer

```{r}
# Load climate layer
climate_flanders <- st_read(file.path(data_path, "klimaatvl.shp")) %>%
  st_transform(crs = 31370)

# Glimpse of data
glimpse(climate_flanders)
```

We calculate get first two PC axes of all climate variables.

```{r}
pca <- prcomp(as.matrix(st_drop_geometry(climate_flanders[-1])), scale. = TRUE)
plot(pca)
```

```{r}
pca$rotation %>%
  as.data.frame() %>%
  rownames_to_column(var = "climVar") %>%
  select(climVar, PC1, PC2) %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point() +
  geom_text(aes(label = climVar), nudge_x = -0.05, nudge_y = 0.02)
```

We also select a subset of the variables as such.

```{r}
clim_vars <- c("MAXT1", "GEMMAXJ", "NEERSLAG1", "GEMNEERSLA", "ZON1",
               "GEMZONJAAR", "ZVORSTDAGE")
```

We select `r clim_vars` variables.

```{r}
for (var in clim_vars) {
  p <- climate_flanders %>%
    ggplot(aes(fill = .data[[var]])) +
    geom_sf() +
    scale_fill_viridis_c()
  print(p)
}
```

We add the variables and first two PCs to the observation data.

```r
#| warning: false
site_centroids <- st_intersection(
  site_centroids,
  climate_flanders %>%
    bind_cols(as.data.frame(pca$x)) %>%
    select(all_of(c(clim_vars, "PC1", "PC2")))
)
```

Visualisation:

```r
site_centroids %>%
  pivot_longer(
    cols = all_of(c(clim_vars, "PC1", "PC2")),
    names_to = "climVar",
    values_to = "climVal"
  ) %>%
  ggplot(aes(y = climVal, x = habClass)) +
  geom_boxplot() +
  labs(x = "") +
  facet_wrap(~climVar, ncol = 3, scales = "free") +
  theme_minimal()
```

## Soil layer

> to do

# Sampling effort

We calculate total number of observations per habitat over all pollinator groups.

```{r}
sampling_effort <- bind_rows(pollinator_obs_filtered) %>%
  st_drop_geometry() %>%
  count(habID, name = "samplingEffort")

plot_data <- left_join(site_centroids, sampling_effort, by = join_by(habID))
```

Visualisation of logarithm of sampling effort:

```{r}
plot_data %>%
  ggplot(aes(y = log(samplingEffort), x = habClass)) +
  geom_boxplot() +
  labs(x = "") +
  theme_minimal()
```

# Write out plot dataset

```{r}
write_csv(
  plot_data %>%
    arrange(habID) %>%
    mutate(
      x = st_coordinates(geometry)[, 1],
      y = st_coordinates(geometry)[, 2]
    ) %>%
    st_drop_geometry(),
  file.path(out_path, "plot_data_wide.csv")
)
```
