---
title: "Data exploration"
author: "Ward Langeraert"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
editor_options: 
  chunk_output_type: console
---

# Goal

Data exploration.

```{r}
#| warning: false
#| message: false
# Packages
library(tidyverse)

# Conflicts
conflicted::conflicts_prefer(dplyr::filter)

# Paths
interim_path <- here::here("data", "intermediate")
data_path <- here::here("data", "processed")
```

# Load data

Load observation data of pollinator groups.

```{r}
poll_groups <- c("bees", "butterflies", "moths", "hoverflies")

# Read the files and build a named list
pollinator_obs_df <- lapply(
  file.path(data_path, paste0(poll_groups, "_data.csv")),
  read_csv,
  show_col_types = FALSE
)
names(pollinator_obs_df) <- poll_groups
```

# Univariate

Number of observations per pollinator group.

```{r}
bind_rows(pollinator_obs_df) %>%
  count(pollinatorGroup) %>%
  ggplot(aes(x = pollinatorGroup, y = n)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n), vjust = -0.25) +
  labs(x = "", y = "Number of observations") +
  theme_minimal()
```

Number of species per pollinator group.

```{r}
bind_rows(pollinator_obs_df) %>%
  group_by(pollinatorGroup) %>%
  summarise(n_spec = n_distinct(species), .groups = "drop") %>%
  ggplot(aes(x = pollinatorGroup, y = n_spec)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n_spec), vjust = -0.25) +
  labs(x = "", y = "Number of species") +
  theme_minimal()
```

We look at the 50 most common observed per group.
We do not account for sampling effort.

```{r}
for (i in names(pollinator_obs_df)) {
  p <- pollinator_obs_df[[i]] %>%
    count(vernacularName) %>%
    slice_max(n, n = 50) %>%
    mutate(vernacularName = reorder(vernacularName, n)) %>%
    ggplot(aes(x = n, y = vernacularName)) +
    geom_bar(stat = "identity") +
    labs(x = "Number of observations", y = "") +
    theme_minimal()

  print(p)
}
```

There are some very rare species here that seem more common than species we know are more common.

We look at species that occur in most UTM1 kilometre squares.
However, some grid cells have been visited more than others.
We divide the number of grid cells per species by the sum of the sampling effort (number of observation for each cells).

```{r}
for (i in names(pollinator_obs_df)) {
  p <- pollinator_obs_df[[i]] %>%
    distinct(vernacularName, utm1, samplingEffort) %>%
    summarise(
      n = n(),
      effort = sum(samplingEffort),
      test = n / effort,
      .by = "vernacularName"
    ) %>%
    slice_max(n, n = 50) %>%
    mutate(vernacularName = reorder(vernacularName, n)) %>%
    ggplot(aes(x = n, y = vernacularName)) +
    geom_bar(stat = "identity") +
    labs(x = "Number of occupied UTM1 squares (sampling effort)", y = "") +
    theme_minimal()

  print(p)
}
```

This looks like a good representation for commonness.

```{r}
common_specs <- lapply(pollinator_obs_df, function(df) {
  df %>%
    distinct(species, utm1, samplingEffort) %>%
    summarise(
      n = n(),
      effort = sum(samplingEffort),
      test = n / effort,
      .by = "species"
    ) %>%
    slice_max(n, n = 50) %>%
    mutate(species = reorder(species, n)) %>%
    pull(species)
})
```

# Bivariate
## Categorical
### Habitat class

We look at the number of observations per habitat class.

```{r}
bind_rows(pollinator_obs_df) %>%
  count(pollinatorGroup, habClass) %>%
  ggplot(aes(x = habClass, y = n, fill = pollinatorGroup)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(y = "Number of observations", x = "") +
  geom_text(aes(label = n), vjust = -0.25,
            position = position_dodge(width = 0.9)) +
  theme_minimal()
```

We look at the number of species per habitat class.

```{r}
bind_rows(pollinator_obs_df) %>%
  group_by(pollinatorGroup, habClass) %>%
  summarise(n_spec = n_distinct(species), .groups = "drop") %>%
  ggplot(aes(x = habClass, y = n_spec, fill = pollinatorGroup)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = n_spec), vjust = -0.25,
            position = position_dodge(width = 0.9)) +
  labs(y = "Number of species", x = "") +
  theme_minimal()
```

### Soil class



## Continuous
### Size of habitat parcels

### Climatic variables
