---
title: "Preparation of observation data"
author: "Ward Langeraert"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
editor_options: 
  chunk_output_type: console
---

# Goal

Load and prepare the observation data in `data/raw` folder.
This includes merging of datasets and use simple filters.

```{r}
#| warning: false
#| message: false
# Packages
library(tidyverse)
library(knitr)

# Conflicts
conflicted::conflicts_prefer(dplyr::filter)

# Paths
source_path <- here::here("source", "scripts")
data_path <- here::here("data", "raw")
out_path <- here::here("data", "intermediate")

# Source
source(file.path(source_path, "get_base.R"))
```

# Load data

```{r}
# Get files with observation data
observation_files <- list.files(data_path, "^INBODATAVR", full.names = TRUE)
observation_files <- observation_files[
  !grepl("MetToestemming", observation_files, ignore.case = TRUE)
]

# Get base names for dataframe objects
bases <- tolower(sapply(observation_files, get_base, USE.NAMES = FALSE))

# Make bases unique when needed (e.g. nachtvlinders_df1, _df2 …)
bases_unique <- make.unique(bases, sep = "_df")  # appends _df1, _df2, etc.

# Read the files and build a named list
insect_dfs <- lapply(observation_files, read_csv, show_col_types = FALSE)
names(insect_dfs) <- bases_unique

# Join datasets of the same groups
idx_groups <- split(seq_along(bases), bases)   # e.g. $nachtvlinders = 3 4 5 6

merged_dfs <- lapply(idx_groups, function(idx) {
  bind_rows(insect_dfs[idx])   # align columns, fill in NA where absent
})
```

Which groups do we have?

```{r}
names(merged_dfs)
```

How many records per group?

```{r}
sapply(merged_dfs, nrow)
```

# Filter data
## Date

We select the data from 2015-2024 (10 year span).

```{r}
# Select data in time frame
merged_dfs <- lapply(merged_dfs, function(df) {
  df %>%
    filter(year(datum) > 2014,
           year(datum) <= 2024)
})
```

## Validated data

We select data that has been validated.

```{r}
# Select validated data
merged_dfs <- lapply(merged_dfs, function(df) {
  df %>%
    filter(grepl("^Goedgekeurd", status))
})
```

## Precision of coordinates

How is the precision distributed? We only look at precision of less then 500 m.

```{r}
#| warning: false
#| message: false
for (i in names(merged_dfs)) {
  p <- merged_dfs[[i]] %>%
    filter(precisie < 500) %>%
    ggplot(aes(x = precisie)) +
    geom_histogram() +
    ggtitle(i)
  print(p)
}
```

We select data that has been validated.

```{r}
# Select data with small precision
merged_dfs <- lapply(merged_dfs, function(df) {
  df %>%
    filter(precisie < 100)
})
```

## Duplicates

Is every record unique?

```{r}
length(unique(unlist(sapply(merged_dfs, function(df) unique(df$id))))) -
  sum(sapply(merged_dfs, nrow))
```

No. There are more observations then unique id's.
Let's check 10 random examples from each group:

```{r}
#| results: "asis"
set.seed(123)

for (i in names(merged_dfs)) {
  double_ids <- merged_dfs[[i]] %>%
    group_by(id) %>%
    filter(n() > 1) %>%
    ungroup()

  ids <- sample(unique(double_ids$id), 10)

  subset_df <- double_ids %>%
    filter(id %in% ids) %>%
    select(id, naam_nl, datum, aantal, gedrag, methode, geslacht, lon, lat)

  print(i)
  cat("\n")
  print(kable(subset_df, digits = 5))
  cat("\n")
}
```

We notice that this is because you can have multiple individuals in a single observation ID (see e.g. [here](https://waarnemingen.be/observation/246980222/) or [here](https://waarnemingen.be/observation/178279003/) or [here](https://waarnemingen.be/observation/194235586/)).
